---
title: 'Logistic Regression Analysis: Pima Indians Diabetes'
author: "Tilak Poudel MDS 3rd sem, Roll no. 36"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    highlight: tango
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

```

# Load Libraries

```{r 'load libraries'}
library(mlbench)   # For PimaIndiansDiabetes2
library(tidyverse) # For data transformation & plots
library(caret)     # For partitioning & confusion matrix
library(ROCR)      # For ROC & AUC calculation
library(mice)      # For handling missing values via imputation
library(car)       # For VIF calculations

```

# 1. Environment Setup & Data Loading
First, we load the required libraries for our analytical pipeline. We use `mlbench` for our source data,`mice` for handling missing features, `tidyverse` for data manipulation/visualization, and `caret/ROCR` for cross-validation diagnostics.

```{r load data}
data("PimaIndiansDiabetes2")
glimpse(PimaIndiansDiabetes2)
summary(PimaIndiansDiabetes2)
```

# 2. Data Cleaning & Imputation
```{r data cleaning}

colSums(is.na(PimaIndiansDiabetes2))

```

To avoid discarding partial information, we execute Multiple Imputation by Chained Equations (MICE) using Predictive Mean Matching (PMM) to fill missing data points based on surrounding patient trends.

```{r imputation}

set.seed(123)
# Run imputation routing
imputed_data <- mice(PimaIndiansDiabetes2, m = 1, method = "pmm", printFlag = FALSE)

# Build out the completed dataset
clean_diabetes <- complete(imputed_data)

sum(is.na(clean_diabetes))

```

# 3. Exploratory Data Analysis (EDA)
Before building our statistical equations, we check the baseline distribution of the target variable (diabetes) to establish class balances.

```{r check data }
prop.table(table(clean_diabetes$diabetes))

```

We now isolate plasma glucose values across positive and negative diagnostic states to visualize group variance.

```{r eda}
ggplot(clean_diabetes, aes(x = diabetes, y = glucose, fill = diabetes)) +
    geom_boxplot(alpha = 0.7) +
    theme_minimal() +
    labs(
      title = "Glucose Concentration by Diabetes Status", 
      x = "Diabetes Status", 
      y = "Plasma Glucose"
    )
```

The boxplot visualizing plasma glucose concentration against diabetes status demonstrates a clear, foundational biological relationship: `individuals diagnosed with diabetes typically exhibit significantly higher blood glucose levels compared to those without the condition`.

### Observation:
The graph provides visual proof that `glucose` is a highly significant differentiator, while simultaneously showing enough variance and overlap to confirm that a multi-variable statistical model is necessary to make accurate predictions.

# 4. Data Splitting

We perform an 80/20 train-test split using stratified sampling via caret to ensure equivalent base-rate metrics in both validation matrices.


```{r data splitting}
set.seed(42)

# Generate partition indexes
train_index <- createDataPartition(clean_diabetes$diabetes, p = 0.8, list = FALSE)

train_set <- clean_diabetes[train_index, ]
test_set  <- clean_diabetes[-train_index, ]
```

# 5. Model Estimation & Diagnostics
We construct a binary generalized linear model using the complete matrix of explanatory features under a logit link function.

```{r model fitting}
diabetes_model <- glm(diabetes ~ ., data = train_set, family = binomial)
summary(diabetes_model)
```

### Observation
#### Statistical Significance: Who Matters?

Let's look at the Pr(>|z|) column (the p-values) and the asterisks (***). 
If a p-value is less than $0.05$, that variable is a statistically significant predictor of diabetes.

Highly Significant Predictors (*** or ):

- glucose ($p < 2\times 10^{-16}$) — By far the strongest predictor.
- pregnant ($p = 6.73\times 10^{-6}$)
- mass (BMI) ($p = 7.23\times 10^{-5}$)
- pedigree (Diabetes pedigree function) ($p = 0.002$)

Not Significant Predictors:

- pressure ($p = 0.633$), 
- triceps ($p = 0.743$), 
- insulin ($p = 0.480$), and 
- age ($p = 0.992$).

#### Insight: Notice that age is not significant here. That doesn't mean age doesn't matter in real life; it means that once we already know a patient's glucose, BMI, and pregnancy history, adding their exact age provides no additional predictive power to the model.

## Collinearity Testing
We calculate the Variance Inflation Factor (VIF) to ensure that inter-variable correlations do not warp our standard error values.

```{r multicollinearity test}
vif(diabetes_model)

```
### Observation: 

No VIF metrics cross above the conservative benchmark threshold ($\text{VIF} > 5$). Structural multicollinearity is not an issue in this context.

## Odds Ratios: 
To evaluate effect sizing intuitively, we convert log-odds metrics to standard risk-based Odds Ratios (OR) alongside their $95\%$ confidence bounds.

```{r odd ratios}
exp(cbind(OR = coef(diabetes_model), confint(diabetes_model)))
```

### Observation:

An Odds Ratio tells how the likelihood of the outcome (having diabetes) changes with a 1-unit increase in that specific predictor variable.

- $\text{OR} = 1.0$: The predictor has no effect on the odds of diabetes.
- $\text{OR} > 1.0$: The predictor increases the odds of diabetes (risk factor).
- $\text{OR} < 1.0$: The predictor decreases the odds of diabetes (protective factor).

**Glucose (OR = 1.0416)**:

For every 1-unit increase in plasma glucose concentration, the odds of a patient developing diabetes increase by $4.16\%$ (calculated as $1.0416 - 1 = 0.0416$).

**Mass / BMI (OR = 1.0910)**:

For every 1-unit increase in BMI, a patient's odds of being diabetic increase by $9.1\%$ ($1.0910 - 1$). If someone's BMI jumps by 10 units, their risk multiplies significantly.

# 6. Out-of-Sample Performance Evaluation
Now we apply our fitted parameters to the untouched validation test set. Predictions are generated as probabilities and converted to factor levels using the traditional $0.5$ classification cutoff threshold.

```{r perf evaluation}

# Calculate continuous risk probabilities
test_set$prob <- predict(diabetes_model, newdata = test_set, type = "response")

# Convert probabilities to factors
test_set$pred_class <- factor(ifelse(test_set$prob > 0.5, "pos", "neg"), levels = c("neg", "pos"))

```

## Confusion Matrix Diagnostics
We calculate predictive indicators (Accuracy, Sensitivity, and Specificity) to evaluate spatial performance.

```{r confusion matrix}
confusionMatrix(test_set$pred_class, test_set$diabetes, positive = "pos")
```

### Observation:

 **Accuracy : 0.7647**

- True Negatives (87): The model correctly predicted 87 healthy patients as neg.

- True Positives (30): The model correctly predicted 30 diabetic patients as pos.

- False Positives (13): The model flagged 13 healthy people as diabetic.

- False Negatives (23): The model cleared 23 diabetic patients, telling them they were healthy.

**Sensitivity (0.5660)**:

Out of all the people who actually have diabetes in test pool, the model successfully caught $56.6\%$ of them.

***Critique***: 
This is the weakest link in the current model. It means we are missing $43.4\%$ of the diabetic patients (the 23 False Negatives), sending them home without treatment.

**Specificity (0.8700)**:

Out of all the truly healthy people, the model correctly identified $87.0\%$ of them. It is highly specific, meaning it rarely panics or triggers false alarms for healthy individuals.

## Discrimination Mapping (ROC & AUC)
Finally, we run a threshold sweep across the true positive/false positive spectrum to establish discrimination power via the Area Under the Curve (AUC).


```{r roc-auc}
# 1. Generate performance tracking profiles
pred_obj <- prediction(test_set$prob, test_set$diabetes)
perf_obj <- performance(pred_obj, measure = "tpr", x.measure = "fpr")

# 2. Extract numeric AUC value dynamically
auc_obj   <- performance(pred_obj, measure = "auc")
auc_value <- round(auc_obj@y.values[[1]], 4)
cat("Final Area Under the Curve (AUC):", auc_value, "\n")

# 3. Plot the base ROC Curve Space
plot(perf_obj, col = "darkgreen", lwd = 2, main = "ROC Curve: Pima Indians Diabetes Model")
abline(a = 0, b = 1, lty = 2, col = "red") # Random baseline

# 4. Corrected Legend Overlay
legend("bottomright", 
       legend = paste("AUC =", auc_value), 
       bty = "o",           # Draws a complete border box ("o" for outer box)
       box.col = "gray",    # Border color
       bg = "white",        # Box background fill
       inset = 0.05)        # Slight padding away from the exact margins

```

# 7. Real-Time Risk Prediction for Sample Patients

We pass completely new, hypothetical patient observations through our model matrix to observe how risk calculations change based on clinical traits.

```{r patient-predictions}
# Create custom sample patients matching the original data framework
new_patients <- tibble(
  pregnant = c(6, 1),
  glucose  = c(165, 95),
  pressure = c(82, 70),
  triceps  = c(35, 18),
  insulin  = c(180, 60),
  mass     = c(38.5, 22.0),
  pedigree = c(1.25, 0.25),
  age      = c(45, 24)
)

# Assign readable profile names
rownames(new_patients) <- c("Patient_1_HighRisk", "Patient_2_LowRisk")

# Predict exact continuous probabilities
new_patients$Predicted_Probability <- predict(diabetes_model, newdata = new_patients, type = "response")

# Convert to final binary decision tag (using our standard 0.5 threshold)
new_patients$Predicted_Class <- ifelse(new_patients$Predicted_Probability > 0.5, "Positive", "Negative")

# Display the prediction matrix
knitr::kable(new_patients %>% select(glucose, mass, pedigree, Predicted_Probability, Predicted_Class), 
             caption = "Model Assessment of New Patient Samples")
