---
title: 'Logistic Regression Analysis: Pima Indians Diabetes'
author: "Data Analysis Report"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    highlight: tango
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

```

# Load Libraries

```{r 'load libraries'}
library(mlbench)   # For PimaIndiansDiabetes2
library(tidyverse) # For data transformation & plots
library(caret)     # For partitioning & confusion matrix
library(ROCR)      # For ROC & AUC calculation
library(mice)      # For handling missing values via imputation
library(car)       # For VIF calculations

```

# 1. Environment Setup & Data Loading
First, we load the required libraries for our analytical pipeline. We use `mlbench` for our source data,`mice` for handling missing features, `tidyverse` for data manipulation/visualization, and `caret/ROCR` for cross-validation diagnostics.

```{r load data}
data("PimaIndiansDiabetes2")
glimpse(PimaIndiansDiabetes2)
```

# 2. Data Cleaning & Imputation
```{r data cleaning}

colSums(is.na(PimaIndiansDiabetes2))

```

To avoid discarding partial information, we execute Multiple Imputation by Chained Equations (MICE) using Predictive Mean Matching (PMM) to fill missing data points based on surrounding patient trends.

```{r imputation}

set.seed(123)
# Run imputation routing
imputed_data <- mice(PimaIndiansDiabetes2, m = 1, method = "pmm", printFlag = FALSE)

# Build out the completed dataset
clean_diabetes <- complete(imputed_data)

sum(is.na(clean_diabetes))

```

# 3. Exploratory Data Analysis (EDA)
Before building our statistical equations, we check the baseline distribution of the target variable (diabetes) to establish class balances.

```{r check data }
prop.table(table(clean_diabetes$diabetes))

```

We now isolate plasma glucose values across positive and negative diagnostic states to visualize group variance.

```{r eda}
ggplot(clean_diabetes, aes(x = diabetes, y = glucose, fill = diabetes)) +
    geom_boxplot(alpha = 0.7) +
    theme_minimal() +
    labs(
      title = "Glucose Concentration by Diabetes Status", 
      x = "Diabetes Status", 
      y = "Plasma Glucose"
    )
```

# 4. Data Splitting

We perform an 80/20 train-test split using stratified sampling via caret to ensure equivalent base-rate metrics in both validation matrices.


```{r data splitting}
set.seed(42)

# Generate partition indexes
train_index <- createDataPartition(clean_diabetes$diabetes, p = 0.8, list = FALSE)

train_set <- clean_diabetes[train_index, ]
test_set  <- clean_diabetes[-train_index, ]
```

# 5. Model Estimation & Diagnostics
We construct a binary generalized linear model using the complete matrix of explanatory features under a logit link function.

```{r model fitting}
diabetes_model <- glm(diabetes ~ ., data = train_set, family = binomial)
summary(diabetes_model)
```

## Collinearity Testing
We calculate the Variance Inflation Factor (VIF) to ensure that inter-variable correlations do not warp our standard error values.

```{r multicollinearity test}
vif(diabetes_model)

```
Note: No VIF metrics cross above the conservative benchmark threshold ($\text{VIF} > 5$). Structural multicollinearity is not an issue in this context.

## Odds Ratios: 
To evaluate effect sizing intuitively, we convert log-odds metrics to standard risk-based Odds Ratios (OR) alongside their $95\%$ confidence bounds.

```{r odd ratios}
exp(cbind(OR = coef(diabetes_model), confint(diabetes_model)))
```
# 6. Out-of-Sample Performance Evaluation
Now we apply our fitted parameters to the untouched validation test set. Predictions are generated as probabilities and converted to factor levels using the traditional $0.5$ classification cutoff threshold.

```{r perf evaluation}

# Calculate continuous risk probabilities
test_set$prob <- predict(diabetes_model, newdata = test_set, type = "response")

# Convert probabilities to factors
test_set$pred_class <- factor(ifelse(test_set$prob > 0.5, "pos", "neg"), levels = c("neg", "pos"))

```

## Confusion Matrix Diagnostics
We calculate predictive indicators (Accuracy, Sensitivity, and Specificity) to evaluate spatial performance.

```{r confusion matrix}
confusionMatrix(test_set$pred_class, test_set$diabetes, positive = "pos")

```

## Discrimination Mapping (ROC & AUC)
Finally, we run a threshold sweep across the true positive/false positive spectrum to establish discrimination power via the Area Under the Curve (AUC).

```{r roc and auc}
# Generate performance tracking profiles
pred_obj <- prediction(test_set$prob, test_set$diabetes)
perf_obj <- performance(pred_obj, measure = "tpr", x.measure = "fpr")

# Plot ROC Space
plot(perf_obj, col = "darkgreen", lwd = 2, main = "ROC Curve: Pima Indians Diabetes Model")
abline(a = 0, b = 1, lty = 2, col = "red") # Random baseline

# Isolate numeric AUC value
auc_obj <- performance(pred_obj, measure = "auc")
cat("Final Area Under the Curve (AUC):", round(auc_obj@y.values[[1]], 4), "\n")

```




